### 2020-12-19 latex/conclusion.tex
Added a better intro, edited symbols and concusion

\section{Limitations and Future Research}
Our VNHC framework is relies on the following assumptions:
\begin{enumerate}
    \item The input matrix \(B(q) \equiv B \in \R^{n \times k}\) is constant and
        full rank.
    \item The input matrix has a left-annihilator 
        \(B^\perp \in \R^{(n-k)\times n}\). 
    \item The annihilator matrix \(B^\perp\) is right semi-orthogonal.
    \item The inertia matrix of the system satisfies 
        \(\nabla_{q_u}M(q) = \Zmat{n(n-k) \times n}\).
    \item On the constraint manifold \(\Gamma\), one can solve for \(q_a\) as a
        function of \((q_u,p_u)\).
\end{enumerate}
If any of these assumptions are not satisfied, one may not be able to find the
constrained dynamics in simply actuated coordinates, and the results may not
hold.
In particular, the theoretical guarantees of this thesis do not hold for systems
with physical nonholonomic constraints like friction, nor does it apply to vehicles
with wheels.

This guides us to the following research directions. 
First, one might consider relaxing the assumptions on the input and inertia
matrices to widen the class of systems to which VNHCs can be applied.
Second, one might forgo the assumption that \(q_a\) is solvable on the
constraint manifold; instead, one could borrow from the VHC literature
and parameterize the constraint manifold with a different set of
coordinates than \((q_u,p_u)\).
Third, one might consider abstract mechanical systems with dynamics
\[
    \begin{cases}
        \dot{q} = \Minv(q) p
        , \\
        \dot{p} = -\pdmat - \nabla_q V(q) + Q_{nh}(q,p) + B(q)\tau
        ,
    \end{cases}
\]
where \(Q_{nh}(q,p) \in \R^n\) are the equations for physical nonholonomic
constraints like friction.
Fourth, one can investigate a smoother mechanism for energy stabilization, where
a VNHC is explicitly designed to inject energy below some energy level, and
dissipate energy above it.
For example, this might involve a mechanism for smoothly transferring between 
VNHCs while maintaining certain safety requirements.

Finally, the proof of Theorem \ref{thm:acrobot-energy-stabilization} 
in Chapter \ref{sec:acrobot-proof} opens a door to the idea of VNHC generation. 
One might search for suitable conditions on a VNHC whereby the Poincar\'{e}
sections for oscillations and rotations are increasing.
Then, one can extend Otsason's ``virtual constraint
generator" \cite{vcg} with the goal of generating regular VNHCs which satisfy
these energy injection/dissipation conditions.
In this way, one could automatically generate VNHCs which stabilize a
desired energy level.

>>>>

\section{Limitations and Future Research}
Our VNHC framework relies on the following assumptions:
\begin{enumerate}
    \item The input matrix \(B(q) \equiv B \in \R^{n \times k}\) is constant and
        full rank.
    \item The input matrix has a left-annihilator 
        \(B^\perp \in \R^{(n-k)\times n}\). 
    \item The annihilator matrix \(B^\perp\) is right semi-orthogonal.
    \item The inertia matrix of the system satisfies 
        \(\nabla_{q_u}M(q) = \Zmat{n(n-k) \times n}\).
    \item On the constraint manifold \(\Gamma\), one can solve for \(q_a\) as a
        function of \((q_u,p_u)\).
\end{enumerate}
If any of these assumptions are not satisfied, one may not be able to find the
constrained dynamics in simply actuated coordinates, and the results of this
thesis may not hold.
In particular, the theoretical guarantees of this thesis do not apply to systems
with physical nonholonomic constraints (\eg friction), nor do they apply to vehicles
with wheels.

These limitations guide us toward the following research directions. 
First, one might relax the assumptions on the input and inertia
matrices, thereby widening the class of systems to which VNHCs can be applied.
Second, one might forgo the assumption that \(q_a\) is solvable on the
constraint manifold; instead, one could borrow from the VHC literature
and parameterize the constraint manifold with a different set of
coordinates than \((q_u,p_u)\).
Third, one might consider more general mechanical systems with dynamics given by
\[
    \begin{cases}
        \dot{q} = \Minv(q) p
        , \\
        \dot{p} = -\pdmat - \nabla_q V(q) + Q_{nh}(q,p) + B(q)\tau
        ,
    \end{cases}
\]
where \(Q_{nh}(q,p) \in \R^n\) are terms corresponding to physical nonholonomic
constraints like friction.
Fourth, one can investigate a smoother mechanism for energy stabilization, where
a VNHC is explicitly designed to inject energy below some energy level, and
dissipate energy above it.
For example, this might involve a mechanism for smoothly transferring between 
VNHCs while maintaining certain safety requirements.

Finally, the proof of Theorem \ref{thm:acrobot-energy-stabilization} 
in Chapter \ref{sec:acrobot-proof} opens a door to the possibility of VNHC generation. 
One might search for suitable conditions on VNHCs where the Poincar\'{e}
sections for oscillations and rotations are increasing.
Then, one can extend Otsason's ``virtual constraint
generator" \cite{vcg} to generate regular VNHCs which satisfy
these energy injection/dissipation conditions.
In this way, one could automatically generate regular VNHCs which stabilize a
desired energy level.

### 2020-12-19 latex/introduction.tex
Added a better intro, edited symbols and concusion

Consider a child on a standing swing.
Naturally, they begin by pushing off the ground and proceed to alternate between
standing and squatting. 
This allows them to swing higher and faster over time, until they have reached
their express goal of going as high and fast as possible.

Now imagine that the child is a robot, and their creator is teaching them to
behave like a real boy.
If the roboticist had studied classical control theory, they would plot the leg
motion of a human child as a trajectory over time and tell the robot to
synchronize with this trajectory.
In an ideal world, this technique would work perfectly; unfortunately, this
control approach is entirely unnatural. 
Humans on swings do not have an internal stopwatch.
Rather, they adjust how much they bend their knees based on the current angle of
the swing along with their direction of motion.
This position-velocity adjustment allows humans to correct for external
disturbances (like strong winds or overly enthusiastic parents).
Figure \ref{fig:swing-pos-vel} shows this usual adjustment:
a person will squat at the peak of the swing and stand as they reach the fastest
point at the bottom \cite{pumping_swing_standing_squatting}.
Even if the robot could perfectly track a time-based trajectory, an external
disturbance may swap the order of squatting and standing to the point that
the swing slows down rather than speeding up.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{images/swing_pos_vel.png}
    \caption{A person on a standing swing will squat as they approach the
        bottom of the swing path, and stand after they pass it until they hit
        the peak of their swing. Image taken and modified from
    \cite{pumping_swing_standing_squatting}.}
    \label{fig:swing-pos-vel}
\end{figure}

One recent control technique known as \textit{virtual holonomic constraints}
\cite{vhcs_for_el_systems} forces a robot's actuators to track a function of
position rather than time.
Sadly, this will not work for the swing-bot: one needs to know their velocity to
determine when they have reached the peak of their swing.
In other words, the robot needs to move according to a function of both position
and velocity if it has any hope of behaving like a real person.
These position-velocity controllers are known as 
\textit{virtual nonholonomic constraints}, and they are the focus of this
thesis.

Throughout this thesis we will refer to two examples to build intuition.
The first is the variable-length pendulum, which models the child on a swing.
The second is the acrobot, which models a gymnast hanging on a horizontal bar.
For each of these mechanical systems, we will develop virtual nonholonomic
constraints which replicate human behaviours. 
Finally, we will rigorously prove that these constraints inject energy into
their respective robots.

\section{Literature Review}
Here we review some of the previous work on energy injection and virtual
nonholonomic constraints.

Energy injection for mechanical systems is often performed by passivity-based
control and energy shaping --- one views the robot and its inputs as energy
transformations which are moulded to force the system to behave as desired.
Much of the theoretical work in this field has been performed by Ortega, van der
Schaft, and others (see for example \cite{ida_pbc_underactuation_one,
ida_pbc_acrobot_example,energy_shaping_revisited}).
Energy shaping has been applied to both the variable-length
pendulum \cite{vlp_energy_shaping} and the acrobot
\cite{swingup_acrobot_energy,swingup_giant_acrobot}.
This technique is useful, but it does not necessarily produce the structured
human-like motion we wish to generate using virtual nonholonomic constraints.

The modern concept of virtual nonholonomic constraints was described by
\citet{nhvc_dynamic_walking} in 2015, though there are references to more
primitive versions going back as early as the the year 2000
\cite{vnhc_human_robot_cooperation}.
Virtual nonholonomic constraints have been of most notable use in the area of
bipedal locomotion \cite{nhvc_incline_walking,output_nhvc_bipedal_conrol},
where they have shown marked improvements in the robustness of walking gaits
when compared to previous control technqiues \cite{nhvc_incline_walking}.
They have also been used for error-reduction in time-delayed teleoperation
\cite{vnhc_time_delay_teleop} and human-robot interaction
\cite{psd_based_vnhc_redundant_manipulator,haptic_vnhc}.
All these fields use different formulations of virtual nonholonomic constraints;
\citet{hybrid_zero_dynamics_bipedal_nhvcs} have derived the constrained dynamics
of virtual nonholonomic constraints for hybrid mechanical systems,
but to the best of our knowledge no one has attempted to generalize the
concept to abstract mechanical systems.

Many of our contributions in Chapter \ref{ch:vnhcs} extend
concepts from the field of virtual \textit{holonomic} constraints. 
Of note is the theoretical framework devised by Mohammadi, Maggiore, and
Consolini, among others
\cite{vhcs_for_el_systems,dynamic_vhcs_stabilize_closed_orbits,lagrangian_structure_reduced_dynamics_vhcs,xingbo_thesis}.

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=0.6\textwidth,angle=270]{images/kiiking.jpg}
    \caption{Kiiking, which translates to ``swing", involves a person standing
        and squatting on a giant swing until they are rotating around the bar.
    Image taken from \cite{kiiking-img}.}
    \label{fig:kiiking}
\end{figure}

The Estonian sport of ``Kiiking" (Figure \ref{fig:kiiking}) is the adult
equivalent of a child playing on a standing swing. 
Here, an athlete attempts to make a giant solid steel swing gain momentum.
Naturally, the kiiker begins by pushing off the ground and proceeds to alternate
between standing and squatting. 
This allows them to swing higher and faster over time, until they have reached
their express goal of spinning at high speed around the bar.

Now imagine that the kiiker is a robot, and their creator is teaching them to
behave like a real boy.
If the roboticist had studied classical control theory, they would plot 
a human's squat depth as a trajectory over time and tell the robot to
synchronize with this trajectory.
In an ideal world, this technique would work perfectly; unfortunately, this
control approach is entirely unnatural. 
After all, humans on swings do not have an internal stopwatch telling them when
to squat.
Rather, they adjust how much they bend their knees based on the current angle of
the swing along with their direction of motion.
This position-velocity adjustment allows humans to correct for external
disturbances (like strong winds or overly enthusiastic pushers). 
Figure \ref{fig:swing-pos-vel} shows this usual adjustment:
a person will squat at the peak of the swing and stand as they reach the fastest
point at the bottom \cite{pumping_swing_standing_squatting}.
Even if the robot could perfectly track a time-based trajectory, an external
disturbance may swap the order of squatting and standing to the point that
the swing can never begin rotating.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{images/swing_pos_vel.png}
    \caption{A person on a standing swing will squat as they approach the
        bottom of the swing path, and stand after they pass it until they hit
        the peak of their swing. Figure modified from
    \cite{pumping_swing_standing_squatting}.}
    \label{fig:swing-pos-vel}
\end{figure}

One recent control technique known as \textit{virtual holonomic constraints}
\cite{vhcs_for_el_systems} forces a robot's actuators to track a function of
position rather than time.
Sadly, this will not work for the swinging robot since one needs to know their
velocity to determine when they have reached the peak of their swing.
In other words, the robot needs to move according to a function of both position
and velocity if it has any hope of behaving like a real person.
These position-velocity controllers are known as 
\textit{virtual nonholonomic constraints}, and they are the focus of this
thesis.

Our goal throughout this thesis is to develop rigorous proof that virtual
nonholonomic constraints are capable of increasing a robot's momentum through a
process known as \textit{energy injection}.
We will use two benchmark robotic systems to build intuition and test
our theory.
The first system is the variable-length pendulum, which models a person
on a swing.
The second is the acrobot, an actuated double pendulum which models a gymnast
hanging on a horizontal bar.
For each of these examples, we will create virtual nonholonomic
constraints which model human motion to make the robot rotate around a bar.
Finally, we will rigorously prove that these constraints inject energy into
their respective robots.

\section{Literature Review}
Let us review some of the previous work on energy injection and virtual
nonholonomic constraints.

Energy injection for mechanical systems is often performed by passivity-based
control and energy shaping --- one views the robot and its inputs as energy
transformations which are moulded to force the system to behave as desired.
Much of the theoretical work in this field has been performed by Ortega, van der
Schaft, and others (see for example \cite{ida_pbc_underactuation_one,
ida_pbc_acrobot_example,energy_shaping_revisited}).
Energy shaping has been applied to both the variable-length
pendulum \cite{vlp_energy_shaping} and the acrobot
\cite{swingup_acrobot_energy,swingup_giant_acrobot}.
This technique is useful and widely applicable, but it does not
produce the structured human-like motion we wish to generate using virtual
nonholonomic constraints.

The modern concept of virtual nonholonomic constraints was described by
\citet{nhvc_dynamic_walking} in 2015, though there are references to more
primitive versions going back as early as the the year 2000
\cite{vnhc_human_robot_cooperation}.
Virtual nonholonomic constraints have been of most notable use in the area of
bipedal locomotion \cite{nhvc_incline_walking,output_nhvc_bipedal_conrol},
where they have shown marked improvements in the robustness of walking gaits
when compared to previous control technqiues \cite{nhvc_incline_walking}.
They have also been used for error-reduction in time-delayed teleoperation
\cite{vnhc_time_delay_teleop} and human-robot interaction
\cite{psd_based_vnhc_redundant_manipulator,haptic_vnhc}.
All these fields use different formulations of virtual nonholonomic constraints;
\citet{hybrid_zero_dynamics_bipedal_nhvcs} have derived the so-called
constrained dynamics for hybrid mechanical systems under virtual nonholonomic
constraints, but to the best of our knowledge no one has attempted to generalize
the concept to abstract mechanical systems.

Many of our contributions are extensions of concepts from the field of virtual
holonomic constraints. 
Of note is the framework devised by Mohammadi, Maggiore, and
Consolini, among others
\cite{vhcs_for_el_systems,dynamic_vhcs_stabilize_closed_orbits,lagrangian_structure_reduced_dynamics_vhcs,xingbo_thesis}.

### 2020-12-19 latex/symbols.tex
Updated \mathbb{S}^1 to \Sone, made Gamma \simeq \SxR instead of eq.
Added a better intro, edited symbols and concusion

% Define a table for the symbols using the booktabs convention
\noindent
\begin{table}[h!]
\resizebox{\textwidth}{!}{%
\centering
\begin{tabular}{ll}
   \toprule
   Symbol & Definition \\
   \midrule
   \(\n\) & The index set \(\{1,\ldots,n\}\) of natural numbers up to \(n\). \\
   \(\R^n\) & Real numbers in \(n\) dimensions. \\
   \(\Rt{T}\) & Real numbers modulo \(T > 0\), with \(\Rt{\infty} = \R\). \\
   \(\mathbb{S}^1\) & The unit circle, equivalent to \(\Rt{2\pi}\). \\
   \(\mathcal{Q}\) & The configuration manifold of a system. \\
   \(C^r(X;Y)\) & The space of \(r\)-times continuously differentiable functions from \(X\) to \(Y\). \\
         & If \(r = \infty\), the space of smooth functions from \(X\) to \(Y\). \\
   \(\R^{n \times m}\) & The space of real-valued matrices with \(n\) rows and \(m\) columns. \\
   \(\Id{n}\) & The \(n \times n\) identity matrix. \\
   \(\Zmat{n \times m}\) & The \(n \times m\) matrix of all zeros. \\
   \(M_i\) & If \(M\) is a vector, the \(i\)th element of \(M\). \\
         & If \(M\) is a matrix, the \(i\)th column of \(M\). \\
   \(M_{i,j}\) & The value of row \(i\), column \(j\) for the matrix \(M\). \\
   \(\dot{x}\) & Derivative of \(x\) with respect to time \(t\). \\
   \(\nabla_v F\) & If \(F\) is \(\R\)-valued, the gradient of \(F\) with respect to \(v\). \\
            & If \(F\) is \(\R^{n\times n}\)-valued and \(v \in \R^m\), the block gradient \(\left[\nabla_{v_1}F\tpose \ldots \nabla_{v_m}F\tpose\right]\tpose \in \R^{nm \times n}\). \\
   \(dF_v\) & Total differential (Jacobian) of \(F\), equivalent to \((\nabla_v F)\tpose\). \\
   \(\Hess F\) & If \(F : \R^n \rightarrow \R\), the \(n \times n\) Hessian matrix of double derivatives of \(F\). \\
               & If \(F : \R^n \rightarrow \R^k\), the block matrix \((\Hess F_1, \ldots, \Hess F_k) \in \R^{n \times nk}\). \\
   \(\partial_v \partial_w F\) & Derivative matrix of \(F : \R^n \times \R^m \rightarrow \R\), with \((i,j)\) element \(\ppdiff{F}{v_i}{w_j}\). \\
   \(\delta_{i,j}\) & The Kronecker delta: 1 if \(i = j\) and 0 otherwise. \\
   \(\otimes\) & The matrix kronecker product. \\ 
   \(\norm{\cdot}\) & The euclidean norm on \(\R^n\). \\
   \bottomrule
\end{tabular}

>>>>

% Define a table for the symbols using the booktabs convention
\noindent
\begin{table}[h!]
\resizebox{\textwidth}{!}{%
\centering
\begin{tabular}{ll}
   \toprule
   Symbol & Definition \\
   \midrule
   \(\R^n\) & Real numbers in \(n\) dimensions. \\
   \(\Rt{T}\) & Real numbers modulo \(T > 0\), with \(\Rt{\infty} := \R\). \\
   \(\mathbb{S}^1\) & The unit circle, diffeomorphic to \(\Rt{2\pi}\). \\
   \(\mathcal{Q}\) & The configuration manifold of a system. \\
   \(C^r(X;Y)\) & The space of \(r\)-times continuously differentiable functions from \(X\) to \(Y\). \\
   \(\R^{n \times m}\) & The space of real-valued matrices with \(n\) rows and \(m\) columns. \\
   \(\Id{n}\) & The \(n \times n\) identity matrix. \\
   \(\Zmat{n \times m}\) & The \(n \times m\) matrix of all zeros. \\
   \(M_i\) & If \(M\) is a vector, the \(i\)th element of \(M\). \\
         & If \(M\) is a matrix, the \(i\)th column of \(M\). \\
   \(M_{i,j}\) & The value of row \(i\), column \(j\) for the matrix \(M\). \\
   \(\dot{x}\) & Derivative of \(x\) with respect to time \(t\). \\
   \(\nabla_v F\) & If \(F\) is \(\R\)-valued, the gradient of \(F\) with respect to \(v\). \\
            & If \(F\) is \(\R^{n\times n}\)-valued and \(v \in \R^m\), the block gradient \(\left[\nabla_{v_1}F\tpose \ldots \nabla_{v_m}F\tpose\right]\tpose \in \R^{nm \times n}\). \\
   \(dF_v\) & Total differential (Jacobian) of \(F\), equivalent to \((\nabla_v F)\tpose\). \\
   \(\partial_w F\) & Partial differential of \(F\) with respect to \(w\), with \(i^\text{th}\) element \(\pdiff{F}{w_i}\). \\
   \(\delta_{i,j}\) & The Kronecker delta: 1 if \(i = j\) and 0 otherwise. \\
   \(\otimes\) & The matrix kronecker product. \\ 
   \(\norm{\cdot}\) & The euclidean norm on \(\R^n\). \\
   \(\simeq\) & Equivalent under diffeomorphism. \\
   \bottomrule
\end{tabular}
